---
title: "Reproducible Research Course Project 2"
Author:
        "Omer Shechter"
Date:
        "24/09/2018"
output: 
  html_document:
    keep_md: true
---

# Synopsis
# Data Processing
### Load the Data Set
Download the files and unzip them into a data directory. 

#### Load R libreries 
```{r results='asis', echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
#The R.Utils is required for the bunzip2 
library(R.utils)
library(dplyr)
```

```{r cache=TRUE}
if(!file.exists("data")) {  
        dir.create("data")  
}  
if(!file.exists("./data/StormData.csv.bz2")) {  
        fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"  
        download.file(fileUrl, destfile="./data/StormData.csv.bz2")  
        if(!file.exists("./data/StormData.csv"))   
                bunzip2("./data/StormData.csv.bz2", "./data/StormData.csv", remove = FALSE, skip = TRUE) 
}
StormsData<-read.csv("./data/StormData.csv")
```

### Initial review  
```{r}
dim(StormsData)
```


using str looking on the dataset 
```{r}
names(StormsData)
```


# Data Processing
Create a subset data frame which will use the relvanet variables 
EVTYPE - The Event type as is 

```{r}
StormDF<-data.frame(Event.Type=StormsData$EVTYPE)
```

The data analyze will be based on years - Create a new column , which will have the begin year of the event
Use the BGN_TIME to extract the years
```{r}
StormDF<-cbind(StormDF,years = with(StormsData,format(as.Date(BGN_DATE, format="%m/%d/%Y"),"%Y")))
```

