---
title: "Reproducible Research Course Project 2"
Author:
        "Omer Shechter"
Date:
        "24/09/2018"
output: 
  html_document:
    keep_md: true
---

# Synopsis
# Data Processing

### Load the Data Set
Download the files and unzip them into a data directory. 

#### Load R libreries 
```{r results='asis', echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
#The R.Utils is required for the bunzip2 
library(R.utils)
library(dplyr)
library(mosaic)
```

```{r}
if(!file.exists("data")) {  
        dir.create("data")  
}  
if(!file.exists("./data/StormData.csv.bz2")) {  
        fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"  
        download.file(fileUrl, destfile="./data/StormData.csv.bz2")  
        if(!file.exists("./data/StormData.csv"))   
                bunzip2("./data/StormData.csv.bz2", "./data/StormData.csv", remove = FALSE, skip = TRUE) 
}
StormsData<-read.csv("./data/StormData.csv")
```

### Initial review  
```{r}
dim(StormsData)
```


using str looking on the dataset 
```{r}
names(StormsData)
```


### Processing
Create a subset data frame which will use the relvanet variables 
EVTYPE - The Event type as is 

```{r}
StormDF<-data.frame(Event.Type=StormsData$EVTYPE)
```

The data analyze will be based on years - Create a new column , which will have the begin year of the event
Use the BGN_TIME to extract the years
```{r}
StormDF<-cbind(StormDF,years = with(StormsData,format(as.Date(BGN_DATE, format="%m/%d/%Y"),"%Y")))
```

Add the Fatalities and injuries columns as is  
```{r}
Fatalities.No<-StormsData$FATALITIES
Injuries.No<-StormsData$INJURIES
StormDF<-cbind(StormDF,data.frame(Fatalities.No))
StormDF<-cbind(StormDF,data.frame(Injuries.No))
```
Add the Damage Columns using the PROPDMG, PROPDMGEXP CROPDMG and the CROPDMGEXP 

Create a Dameg.cost function , based on the values from PROPDMG multiple by 
Diffrent factors (K - 1000 , M - 1000000 , B - 1000000000)
Any other value assume error and put zero 
```{r}
#Copy the Two Damage Columns to the new DataFrame
StormDF$Damage.Cost<-StormsData$PROPDMG
StormDF$Damage.Factor<-StormsData$PROPDMGEXP
# Build a new Column with (K - 1000 , M - 1000000 , B - 1000000000)
#Any other value assume error and put zero 
StormDF<-StormDF %>%  
      mutate(Damage.Cost.Factor = as.numeric(case_when(
      Damage.Factor =="K" ~ 1000,  Damage.Factor =="M" ~ 1000000,
      Damage.Factor =="B" ~ 1000000000,(Damage.Factor !="M" |Damage.Factor !="B" | Damage.Factor !="K")~0)))
#REplace the Damage cost column with the the Damage.Cost * Damage.Cost.Factor
StormDF$Damage.Cost<-StormDF$Damage.Cost * StormDF$Damage.Cost.Factor
#view how table of the cost factor 
table(StormDF$Damage.Cost.Factor)
#remove the temporary columns
StormDF<-select(StormDF,Event.Type,years,Fatalities.No,Injuries.No,Damage.Cost)
head(StormDF)
```

